---
title: "Survival Analysis"
author: "Dr MOHD KHAIRUL ANWAR KAMDI"
date: "11/24/2019"
output: html_document
---

#library
```{r}
library(tidyverse)
library(survival)
library(haven)
library(lubridate)
library(ggplot2)
library(survival) 
library(survminer)
```

#import
```{r}
whas500ds <- read_spss("whas500.sav")
```

#mutate
```{r}
whas <- whas500ds %>% mutate(gender1 = factor(gender, labels = c("male","female")),
                             miord1 = factor(miord, labels= c("first", "recurrent")),
                             mitype1 = factor(mitype, labels=c ("non Q-wave" , "Q-wave")),
                             event = factor(fstat, labels=c("Alive","Dead")))
whas
```

#Select
```{r}
whas2 <- whas %>% select(id,event,admitdate,fdate,age,bmi,sysbp,gender1,miord1,mitype1)
whas2
```

#Descriptive study

```{r}
summary(whas2)
```

Most of the sample was dead, male, 1st MI event and non Q-wave. Mean age was 69.85 years old, mean BMI WAS 26.61 kg/m2, mean systolic BP was 144.7mmHg

#calculate duration
```{r}
whas2$dur <- as.duration(whas2$admitdate %--% whas2$fdate)
whas2$dur_year <- whas2$dur / dyears(1)
summary(whas2)
```

#Visualisation

```{r}
ggplot(whas2, aes(x=factor(event), fill=factor(event))) + 
  geom_bar()
ggplot(whas2, aes(x=factor(event), fill=factor(event))) + 
  geom_bar()+
  facet_wrap(~gender1)
ggplot(whas2, aes(x=factor(gender1), fill=factor(gender1))) + 
  geom_bar()+
  facet_wrap(~event)

```

#Univariable analysis -overall

## kaplan meier estimate

Overall
```{r}
overall_km <-survfit(Surv(time=dur_year,event=="Dead")~1,
                     data=whas2,
                     type="kaplan-meier")
summary(overall_km)
```

## KM plot

```{r}
ggsurvplot(overall_km, data=whas2,risk.table=T)
```

## median surv time

```{r}
quantile(overall_km, probs = 0.5)
```


# Univariable Analysis-gender

##KM Estimate

gender
```{r}
gender_km <-survfit(Surv(time=dur_year,event=="Dead")~gender1,
                     data=whas2,
                     type="kaplan-meier")
summary(gender_km)
```

##KM Plot

```{r}
ggsurvplot(gender_km, data=whas2,risk.table=T, linetype = c(1,2), pval = 1)
```


##Median Survival Time

```{r}
quantile(gender_km, probs = 0.5)
```

Median survival time for male = 5.92 years (95% CI=4.58, NA) while median survival time for female was = 3.61 years (95% CI=2.48, 4.46)


