---
title: "SurvivalAnalysisPrac"
author: "Mohd Azmi"
date: "15/11/2019"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Library

```{r "Library", message=FALSE}
library(haven)
library(dplyr)
library(psych)
library(lubridate)
library(knitr)
library(ggplot2)
library(survival)
library(broom)
library(survminer)
```

## Import Dataset

```{r "Dataset", message=FALSE}
DataSet <- read_dta("stroke_outcome.dta")
```

### Revert Categorical Data

```{r, message=FALSE}
DataSet <- DataSet %>%
  mutate_if(is.labelled, funs(as_factor(.)))
kable(head(DataSet))
```


```{r "Dataset", message=FALSE}
```

alternatively

```{r}
DataSet2 <- read_dta("stroke_outcome.dta")
DataSet2 <- DataSet %>%
  mutate(sex = as_factor(sex),
         race = as_factor(race),
         event = as_factor(event),
         dm2 = as_factor(dm2),
         hpt2 = as_factor(hpt2),
         race2 = as_factor(race2),
         event_s = as_factor(event_s))
kable(head(DataSet2))
```

### Recode Numerical Data to Categorical Data

- SBP
  - explore SBP (to set cut off point)

```{r}
summary(DataSet$sbp)
```

  - set sbp cut off as <= 140, 141 - 160, >160
  
```{r}
DataSet <- DataSet %>%
  mutate(SBP_Cat = cut(DataSet$sbp,
                       c(0,140,160,300),
                       labels = c("140 and below", "141-160", "161 and above")))
```
  


# Descriptive Analysis and Data Exploration

## Summary

```{r , message=FALSE}
kable(summary(DataSet))
kable(describe(DataSet))
```

count duration between 2 date

```{r}
DataSet$Dur_Days <- as.duration(DataSet$doa %--% DataSet$dod) / ddays(1)
kable(head(DataSet))
```

group data by event, and descriptive analysis

- numerical data

```{r}
DataSet %>%
  group_by(event) %>%
  summarise(Mean_Age = mean(age), SD_Age = sd(age),
            Mean_GCS = mean(gcs), SD_GCS = sd(gcs),
            Mean_SBP = mean(sbp), SD_SBP = sd(age),
            Mean_DBP = mean(dbp), SD_DBP = sd(dbp),
            ) %>%
  kable()
```


- categorical data
```{r}
DataSet %>%
  count(event, sex) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~sex)

DataSet %>%
  count(event, race) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~race)

DataSet %>%
  count(event, dm2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~dm2)

DataSet %>%
  count(event, hpt2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~hpt2)

DataSet %>%
  count(event, race2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~race2)

```


- recategorized numerical data

```{r}
DataSet %>%
  count(event, SBP_Cat) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~SBP_Cat)
```


# Univariable

## Overall Model

### Kaplan-Meier Survival Estimates


```{r}
Overall_KM <- survfit(Surv(time = Dur_Days, event == "dead") ~ 1,
                      data = DataSet,
                      type = "kaplan-meier")
summary(Overall_KM)
kable(tidy(Overall_KM))
```

### KM Plot

```{r}
ggsurvplot(Overall_KM,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

## Univariable - IV - gender

### Cox PH Estimate

```{r}
#CPH_gender <- coxph(Surv(time = Dur_Days, event == "dead") ~ sex,
#                     data = DataSet)
#summary(CPH_gender)
#kable(tidy(CPH_gender))
```


### KM Estimate

```{r}
KM_gender <- survfit(Surv(time = Dur_Days, event == "dead") ~ sex,
                     data = DataSet,
                     type = "kaplan-meier")
summary(KM_gender)
kable(tidy(KM_gender))
```

### KM Plot

```{r}
ggsurvplot(KM_gender,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

p = 0.41, thus no significant difference between survival in male or female.

## Univariable - IV - DM

### KM Estimate

```{r}
KM_DM <- survfit(Surv(time = Dur_Days, event == "dead") ~ dm2,
                 data = DataSet,
                 type = "kaplan-meier")
summary(KM_DM)
kable(tidy(KM_DM))
```

### KM Plot

```{r}
ggsurvplot(KM_DM,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

p < 0.05, thus there is significant different between DM yes and DM no

### Survival Function Estimate

estimate number of survival at Q1, median and Q3


```{r}
kable(quantile(KM_DM, probs = c(0.25, 0.5, 0.75)))
```

### Survival Probability Estimate

estimate time of survival at times = 20, 40 and 60 units of time

```{r}
summary(KM_DM, times = c(20, 40, 60))
```

### Compare Survival Estimate - Log Rank Test

```{r}
LogRank_DM <- survdiff(Surv(time = Dur_Days, event == "dead") ~ dm2,
                       data = DataSet,
                       rho = 0)
LogRank_DM
kable(tidy(LogRank_DM))
```

p < 0.05, there is significant different between dm2 yes and dm2 no

### Compare Survival Estimate - Peto-peto

```{r}
Peto_DM <- survdiff(Surv(time = Dur_Days, event == "dead") ~ dm2,
                       data = DataSet,
                       rho = 1)
Peto_DM
kable(tidy(Peto_DM))
```

p < 0.05, there is significant different between dm2 yes and dm2 no


## Univariable - IV - SBP (Category)

## Cox


### KM Estimate

```{r}
KM_SBPCat <- survfit(Surv(time = Dur_Days, event == "dead") ~ SBP_Cat,
                 data = DataSet,
                 type = "kaplan-meier")
summary(KM_SBPCat)
kable(tidy(KM_SBPCat))
```

### KM Plot

```{r}
ggsurvplot(KM_SBPCat,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2,3),
           pval = T)
```

p < 0.05, there is significant differenct between survival rate for SBP (Cat)

## Univariable - IV - GCS (Numerical)

### Cox PH Estimate

Cox PH Estimate with efron method

```{r}
#CPH_GCS <- coxph(Surv(time = Dur_Days,
#                      event == "dead") ~ gcs,
#                 data = DataSet)
#summary(CPH_GCS)
#kable(tidy(CPH_GCS))
```

### KM Estimate

```{r}
KM_GCS <- survfit(Surv(time = Dur_Days,
                       event == "dead") ~ gcs,
                  data = DataSet)
summary(KM_GCS)
kable(tidy(KM_GCS))
```



## Univariable - IV - SBP (Numerical)

### Cox PH Estimate

```{r}
#CPH_SBP <- coxph(Surv(time = Dur_Days,
#                      event == "dead") ~ sbp,
#                 data = DataSet)
#summary(CPH_SBP)
#kable(tidy(CPH_SBP))
```



# Multivariable Analysis

## Predictor - IV - gcs, age, dm, hpt 

### Cox PH 

```{r}
CPH_GCSAgeDMHPT <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2,
                         data = DataSet)
summary(CPH_GCSAgeDMHPT)
kable(tidy(CPH_GCSAgeDMHPT))
```


### Interaction gcs:age

```{r}
CPH_GCSAgeDMHPT_IntGCSAge <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2 + gcs:age,
                         data = DataSet)
summary(CPH_GCSAgeDMHPT_IntGCSAge)
kable(tidy(CPH_GCSAgeDMHPT_IntGCSAge))
```

### Interaction dm:dpt

- model with interaction
```{r}
CPH_GCSAgeDMHPT_IntDMHpt <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2 + dm2:hpt2,
                         data = DataSet)
summary(CPH_GCSAgeDMHPT_IntDMHpt)
kable(tidy(CPH_GCSAgeDMHPT_IntDMHpt))
```

- compare model with interaction using anova

```{r}
anova(CPH_GCSAgeDMHPT_IntDMHpt, CPH_GCSAgeDMHPT, test = "chisq")
```





