---
title: "SurvivalAnalysisPrac"
author: "Mohd Azmi"
date: "15/11/2019"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Library

```{r "Library", message=FALSE}
library(haven)
library(dplyr)
library(psych)
library(lubridate)
library(knitr)
library(ggplot2)
library(survival)
library(broom)
library(survminer)
library(mfp)
```

## Import Dataset

```{r "Dataset", message=FALSE}
DataSet <- read_dta("stroke_outcome.dta")
```

### Revert Categorical Data

```{r, message=FALSE}
DataSet <- DataSet %>%
  mutate_if(is.labelled, funs(as_factor(.)))
kable(head(DataSet))
```


```{r "Dataset", message=FALSE}
```

alternatively

```{r}
DataSet2 <- read_dta("stroke_outcome.dta")
DataSet2 <- DataSet2 %>%
  mutate(sex = as_factor(sex),
         race = as_factor(race),
         event = as_factor(event),
         dm2 = as_factor(dm2),
         hpt2 = as_factor(hpt2),
         race2 = as_factor(race2),
         event_s = as_factor(event_s))
kable(head(DataSet2))
```

### Recode Numerical Data to Categorical Data

- SBP
  - explore SBP (to set cut off point)

```{r}
summary(DataSet$sbp)
```

  - set sbp cut off as <= 140, 141 - 160, >160
  
```{r}
DataSet <- DataSet %>%
  mutate(SBP_Cat = cut(DataSet$sbp,
                       c(0,140,160,300),
                       labels = c("140 and below", "141-160", "161 and above")))
```
  


# Descriptive Analysis and Data Exploration

## Summary

```{r , message=FALSE}
kable(summary(DataSet))
kable(describe(DataSet))
```

## Calculate Duration

count duration between 2 date

```{r}
DataSet$Dur_Days <- as.duration(DataSet$doa %--% DataSet$dod) / ddays(1)
kable(head(DataSet))
```

group data by event, and descriptive analysis

- numerical data

```{r}
DataSet %>%
  group_by(event) %>%
  summarise(Mean_Age = mean(age), SD_Age = sd(age),
            Mean_GCS = mean(gcs), SD_GCS = sd(gcs),
            Mean_SBP = mean(sbp), SD_SBP = sd(age),
            Mean_DBP = mean(dbp), SD_DBP = sd(dbp),
            ) %>%
  kable()
```


- categorical data
```{r}
DataSet %>%
  count(event, sex) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~sex)

DataSet %>%
  count(event, race) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~race)

DataSet %>%
  count(event, dm2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~dm2)

DataSet %>%
  count(event, hpt2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~hpt2)

DataSet %>%
  count(event, race2) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~race2)

DataSet %>%
  count(event, SBP_Cat) %>%
  kable()

ggplot(DataSet, aes(x=factor(event), fill=factor(event))) +
  geom_bar() +
  facet_wrap(~SBP_Cat)
```


# Univariable

## Overall Model

### Kaplan-Meier Survival Estimates


```{r}
Overall_KM <- survfit(Surv(time = Dur_Days, event == "dead") ~ 1,
                      data = DataSet,
                      type = "kaplan-meier")
summary(Overall_KM)
kable(tidy(Overall_KM))
```

### KM Plot

```{r}
ggsurvplot(Overall_KM,
           data = DataSet,
           risk.table = T)
```

### Estimate Survival Function

*Find Duration from given Probability*

Estimate the survival duration at 25^th^, median and 75^th^ percentile

```{r}
kable(quantile(Overall_KM, probs = c(0.25, 0.5, 0.75)))
```

to understand this result, compare with KM Survival Estimate. For example, to estimate the survival duration at median,

- refer KM Survival Estimate, find the duration at survival probability = 0.5
  - at survival probability = 0.5, the duration was 22 days
- similarly, to find the duration at upper limit and lower limit, find the duration at survival probability upper limit 0.5 and lower limit 0.5
  - at survival probability at lower limit = 0.5, the duration was 12 days
  - there was however no survival probability at lower limit = 0.5

*Find Probability from given Duration*

Estimate the survival probability at duration 20 days, 40 days and 60 days

```{r}
summary(Overall_KM, times = c(20,40,60))
```



## Univariable - IV - gender

### KM Estimate

```{r}
KM_gender <- survfit(Surv(time = Dur_Days, event == "dead") ~ sex,
                     data = DataSet,
                     type = "kaplan-meier")
summary(KM_gender)
kable(tidy(KM_gender))
```

### KM Plot

```{r}
ggsurvplot(KM_gender,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

p = 0.41, thus no significant difference between survival in male or female.

### Log Rank Test

to compare survival estimate between male and female, log rank test can be use

```{r}
LR_gender <- survdiff(Surv(time = Dur_Days, event =="dead") ~ sex,
                      data = DataSet,
                      rho = 0)
LR_gender
```

note that the p-value from log rank test is equal to p-value in KM plot. it is because the KM Plot compare using log rank test!



### Cox PH Estimate

```{r}
CPH_gender <- coxph(Surv(time = Dur_Days, event == "dead") ~ sex,
                     data = DataSet)
CPH_gender
summary(CPH_gender)
kable(tidy(CPH_gender))
```


## Univariable - IV - DM

### KM Estimate

```{r}
KM_DM <- survfit(Surv(time = Dur_Days, event == "dead") ~ dm2,
                 data = DataSet,
                 type = "kaplan-meier")
summary(KM_DM)
kable(tidy(KM_DM))
```

### KM Plot

```{r}
ggsurvplot(KM_DM,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

p < 0.05, thus there is significant different between DM yes and DM no

### Survival Function Estimate

estimate number of survival at Q1, median and Q3


```{r}
kable(quantile(KM_DM, probs = c(0.25, 0.5, 0.75)))
```

### Survival Probability Estimate

estimate time of survival at times = 20, 40 and 60 units of time

```{r}
summary(KM_DM, times = c(20, 40, 60))
```

### Compare Survival Estimate - Log Rank Test

```{r}
LogRank_DM <- survdiff(Surv(time = Dur_Days, event == "dead") ~ dm2,
                       data = DataSet,
                       rho = 0)
LogRank_DM
kable(tidy(LogRank_DM))
```

p < 0.05, there is significant different between dm2 yes and dm2 no

### Compare Survival Estimate - Peto-peto

```{r}
Peto_DM <- survdiff(Surv(time = Dur_Days, event == "dead") ~ dm2,
                       data = DataSet,
                       rho = 1)
Peto_DM
kable(tidy(Peto_DM))
```

p < 0.05, there is significant different between dm2 yes and dm2 no


## Univariable - IV - SBP (Category)

### Cox


### KM Estimate

```{r}
KM_SBPCat <- survfit(Surv(time = Dur_Days, event == "dead") ~ SBP_Cat,
                 data = DataSet,
                 type = "kaplan-meier")
summary(KM_SBPCat)
kable(tidy(KM_SBPCat))
```

### KM Plot

```{r}
ggsurvplot(KM_SBPCat,
           data = DataSet,
           risk.table = T,
           linetype = c(1,2,3),
           pval = T)
```

p < 0.05, there is significant differenct between survival rate for SBP (Cat)

## Univariable - IV - GCS (Numerical)

### Cox PH Estimate

Cox PH Estimate with efron method

```{r}
CPH_GCS <- coxph(Surv(time = Dur_Days,
                      event == "dead") ~ gcs,
                 data = DataSet)
CPH_GCS
summary(CPH_GCS)
kable(tidy(CPH_GCS))
```

### KM Estimate

```{r}
KM_GCS <- survfit(Surv(time = Dur_Days,
                       event == "dead") ~ gcs,
                  data = DataSet)
summary(KM_GCS)
kable(tidy(KM_GCS))
```



## Univariable - IV - SBP (Numerical)

### Cox PH Estimate

```{r}
CPH_SBP <- coxph(Surv(time = Dur_Days,
                      event == "dead") ~ sbp,
                 data = DataSet)
CPH_SBP
summary(CPH_SBP)
kable(tidy(CPH_SBP))
```



# Multivariable Analysis

## Predictor - IV - gcs, age, dm, hpt 

### Cox PH 

```{r}
CPH_GCSAgeDMHPT <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2,
                         data = DataSet)
CPH_GCSAgeDMHPT
summary(CPH_GCSAgeDMHPT)
kable(tidy(CPH_GCSAgeDMHPT))
```


### Interaction gcs:age

```{r}
CPH_GCSAgeDMHPT_IntGCSAge <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2 + gcs:age,
                         data = DataSet)
CPH_GCSAgeDMHPT_IntGCSAge
summary(CPH_GCSAgeDMHPT_IntGCSAge)
kable(tidy(CPH_GCSAgeDMHPT_IntGCSAge))
```

### Interaction dm:dpt

- model with interaction
```{r}
CPH_GCSAgeDMHPT_IntDMHpt <- coxph(Surv(time = Dur_Days,
                              event == "dead") ~ gcs + age + dm2 + hpt2 + dm2:hpt2,
                         data = DataSet)
CPH_GCSAgeDMHPT_IntDMHpt
summary(CPH_GCSAgeDMHPT_IntDMHpt)
kable(tidy(CPH_GCSAgeDMHPT_IntDMHpt))
```

- compare model with interaction using anova

```{r}
anova(CPH_GCSAgeDMHPT_IntDMHpt, CPH_GCSAgeDMHPT, test = "chisq")
```

interaction term not significant

# Final Model

## Prelim Final Model

predictors: gcs, age, dm, hpt

### Model Assumption Checking - Linearity in Hazard Regression Assumption

```{r}
mfp_PrelimFinalModel <- mfp(Surv(time = Dur_Days,
                                 event = event == "dead") ~ fp(gcs, df = 4) + fp(age) + dm2 + hpt2,
                            data = DataSet,
                            family = cox,
                            verbose = T)
```

```{r}
mfp_PrelimFinalModel
```

```{r}
head(mfp_PrelimFinalModel$X)
mfp_PrelimFinalModel$powers
mfp_PrelimFinalModel$pvalues
```

### Model Assumption Checking - Proportionality

- KM Method
```{r}
CPH_PrelimFinalModel <- CPH_GCSAgeDMHPT
PropKM_PrelimFinalModel <- cox.zph(CPH_PrelimFinalModel,
                                 transform = "km",
                                 global = T)
PropKM_PrelimFinalModel
plot(PropKM_PrelimFinalModel)
```

- rank method
```{r}
PropRank_PrelimFinalModel <- cox.zph(CPH_PrelimFinalModel,
                                 transform = "rank",
                                 global = T)
PropRank_PrelimFinalModel
plot(PropRank_PrelimFinalModel)
```

- log method
```{r}
PropLog_PrelimFinalModel <- cox.zph(CPH_PrelimFinalModel,
                                 transform = "log",
                                 global = T)
PropLog_PrelimFinalModel
#plot(PropLog_PrelimFinalModel)
```

### Model Checking - Residual (Score)

- score residual
```{r}
ScoreResid_PrelimFinalModel <- resid(CPH_PrelimFinalModel,
                                     type = "score")
kable(head(ScoreResid_PrelimFinalModel))
```

- residual plot (score residual) for GCS (numerical)
```{r}
plot(DataSet$gcs, ScoreResid_PrelimFinalModel[,2], ylab = "Score Residual")
```

- residual plot (score residual) for DM (categorical)
```{r}
plot(DataSet$dm2, ScoreResid_PrelimFinalModel[,2], ylab = "Score Residual")
plot(DataSet$dm2, ScoreResid_PrelimFinalModel[,1], ylab = "Score Residual")
plot(DataSet$dm2, ScoreResid_PrelimFinalModel, ylab = "Score Residual")
```

### Model Checking - Residual (Martingale)

- martingale residual

```{r}
MartinResid_PrelimFinalModel <- resid(CPH_PrelimFinalModel,
                                      type = "martingale")
kable(head(MartinResid_PrelimFinalModel))
```

- residual plot for martingale

```{r}
plot(DataSet$age,
     MartinResid_PrelimFinalModel,
     ylab = "Martingale Residual for Age")
plot(DataSet$sex,
     MartinResid_PrelimFinalModel,
     ylab = "Martingale Residual for Gender")
plot(MartinResid_PrelimFinalModel,
     type = "h",
     main = "Martingale Residuals",
     ylab = "dfbetas",
     lwd = 2)
```

### Martingale Residual - Alternative

```{r}
augment(CPH_PrelimFinalModel,
        data = DataSet)
```

calculate fitted (predicted), and martingale resid


### Model Checking - Residual (Schoenfeld)

- schoenfeld residual
```{r}
SchoenResid_PrelimFinalModel <- resid(CPH_PrelimFinalModel,
                                      type = "schoenfeld")
kable(head(SchoenResid_PrelimFinalModel))
```


### Model Checking - Residual (Scaled Schoenfeld)

- scaled schoenfeld residual

```{r}
SSResid_PrelimFinalModel <- resid(CPH_PrelimFinalModel,
                                  type = "scaledsch")
kable(head(SSResid_PrelimFinalModel))
```

### Model Checking - Residual (dfbeta)

- dfbeta residual

```{r}
dfbResid_PrelimFinalModel <- resid(CPH_PrelimFinalModel,
                                   type = "dfbeta")
kable(head(dfbResid_PrelimFinalModel))
```

- plot for dfbeta

```{r}
plot(DataSet$age,
     dfbResid_PrelimFinalModel[,2],
     main = "Dfbetas for Age",
     ylab = "dfbetas")
plot(DataSet$gcs,
     dfbResid_PrelimFinalModel[,2],
     type = "h",
     main = "Dfbetas for GCS",
     ylab = "dfbetas")
```



# Prediction

## Create New Data

```{r}
NewDS <- expand.grid(gcs = c(mean(DataSet$gcs)),
                     age = c(40, 50, 60),
                     dm2 = c("no", "yes"),
                     hpt2 = c("no", "yes"))
kable(NewDS)
```


## Average Value for All Predictor

## Predict with LP

calculate predicted estimate for data from new dataset

```{r}
kable(predict(CPH_PrelimFinalModel,
        newdata = NewDS,
        type = "lp"))
```

alternatively, but didn't separate from the dataset

```{r}
kable(augment(CPH_PrelimFinalModel,
              newdata = NewDS))
```

## Predict with Risk Score

```{r}
kable(predict(CPH_PrelimFinalModel,
        newdata = NewDS,
        type = "risk"))
```

alternatively, but didnt separate from the dataset

```{r}
kable(augment(CPH_PrelimFinalModel,
              newdata = NewDS,
              type.predict = "risk"))
```

## Another Prediction

add event = dead, and follow up time as 20, 40 and 50 days

```{r}

NewDS2 <- expand.grid(event = 'dead', Dur_Days = c(20, 40, 50))
NewDS3 <- data.frame(NewDS, NewDS2)
```

```{r}
Predicted_Expected <- predict(CPH_PrelimFinalModel,
                              newdata = NewDS3,
                              type = "expected")
kable(cbind(NewDS3, Predicted_Expected))
```

## Predict Survival Probability

to predict survival probability, need to calculate hazard function and linear predictor

$h(T) = h_0(t) * exp^(^/b^X)$


- linear predictor function ($exp^(^/b^x)$)

```{r}
LinearPredFunc <- predict(CPH_PrelimFinalModel,
                          newdata = NewDS,
                          type = "lp")
```

- base hazard function ($h_0(t)$)

```{r}
BaseHF <- basehaz(CPH_PrelimFinalModel)
```

to predict hazard at time 14 days, refer base hazard function at day 14

```{r}
kable(BaseHF)
```

at day 14, base hazard function = 0.3067

to calculated predicted value, times the linear predictor funtion with base hazard function

```{r}
Pred_Value <- 0.3067*exp(LinearPredFunc)
kable(Pred_Value)
```


