---
title: "Survival Analysis Practical"
author: "Mohd Azmi, Mohd Khairul Anwar, Mohd Nasrullah, Mohd Hazwan"
date: "15/11/2019"
output:
    html_document:
      number_sections: true
      toc: yes
      toc_float:
        collapsed: no
---

# Preface

## Research Problem

The dataset taken from Worcester Heart Attack Study. The dataset contain 500 samples, to study factors and time trends associated with long-term survival following acute myocardial infarction.

In this study, we would like to ascertain factors associated with survival following acute myocardial infarction, and predict survival probability.

## Library

```{r}
library(pacman)
p_load(haven, dplyr, psych, lubridate, knitr, ggplot2, survival, broom, survminer, mfp)
```

## Data Wrangling

Import dataset
```{r}
WHASDS <- read_spss("whas500.sav")
WHASDS
```

Convert to categorical data
```{r}
WHASDS2 <- WHASDS %>%
  mutate(gender = factor(gender,
                         labels = c("male", "female")),
         MIOrder = factor(miord,
                          labels = c("first", "recurrent")),
         MIType = factor(mitype,
                         labels = c("non Q-wave", "Q-wave")),
         FUpStatus = factor(fstat,
                            labels = c("alive", "dead"))) %>%
  select (id,
          FUpStatus,
          admitdate,
          fdate,
          lenfol,
          age,
          gender,
          bmi,
          sysbp,
          MIOrder,
          MIType)
WHASDS2
```


calculate duration
```{r}
WHASDS2 <- WHASDS2 %>%
  mutate(dur = as.duration(WHASDS2$admitdate %--% WHASDS2$fdate),
         dur_day = dur/ddays(1),
         dur_year = dur/dyears(1)) %>%
  select (id,
          FUpStatus,
          dur_year,
          age,
          gender,
          bmi,
          sysbp,
          MIOrder,
          MIType,
          admitdate,
          fdate,
          dur_day,
          lenfol)
WHASDS2
```


# Descriptive Study

## Summary

Most of the participant was gender (n=300), first MI episode (n=329) and non Q-wave type (n=347). 
The mean age of the participant was 69.85 years old, with mean BMI 26.61 m/kg^2^, and mean systolic BP of 145 mmHg.

```{r}
WHASDS2 %>%
  select_if(is.numeric) %>%
  summary() %>%
  kable()
WHASDS2 %>%
  select_if(is.factor) %>%
  summary() %>%
  kable()
```

```{r}
ggplot(WHASDS2, aes(x=FUpStatus, fill=FUpStatus)) +
  geom_bar()
ggplot(WHASDS2, aes(x=gender, fill=gender)) +
  geom_bar()
ggplot(WHASDS2, aes(x=MIOrder, fill=MIOrder)) +
  geom_bar()
ggplot(WHASDS2, aes(x=MIType, fill=MIType)) +
  geom_bar()
```

```{r}
ggplot(WHASDS2, aes(age)) +
  geom_histogram(binwidth=2, colour="black", fill="white")
ggplot(WHASDS2, aes(sysbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white")
ggplot(WHASDS2, aes(bmi)) +
  geom_histogram(binwidth=1, colour="black", fill="white")
```


Barchart for all categorical data, divide by follow up status

```{r}
ggplot(WHASDS2, aes(x=gender, fill=gender)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=MIOrder, fill=MIOrder)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=MIType, fill=MIType)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
```

histogram for all numerical data, divide by follow up status
**to add normal curve**
```{r}
ggplot(WHASDS2, aes(age)) +
  geom_histogram(binwidth=2.5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(sysbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(bmi)) +
  geom_histogram(binwidth=1, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
```
 
## KM Estimate For Overall Participants

The overall median survival time was 4.46 years (95% CI = 4.18, NA).

```{r}
Overall_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~1,
                      data = WHASDS2,
                      type = "kaplan-meier")
Overall_KM
```

## KM Plot for Overall Participants

The KM plot for overall participant is shown below.

```{r}
ggsurvplot(Overall_KM,
           data = WHASDS2,
           risk.table = T)
```


## Survival Probability

The 5-year survival probability was 49.4% (95% CI = 44.1%, 55.3%). Other survival probability (for 6 months, 1 year and 3 years) was shown below.

```{r}
summary(Overall_KM, times = c(0.5, 1, 3, 5))
```


# Univariable

## Univariable - Age

### Cox Proportional Hazard

For each 1 year increase of age, the log hazard to die changes by a factor of 0.07 (95% 0.05, 0.08, p < 0.001).

```{r}
age_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~age,
                 data = WHASDS2)
tidy(age_CPH)
tidy(exponentiate = T, age_CPH)
```

## Univariable - Gender

### KM Estimate

The median survival time for male patient was 5.92 years (95% CI = 4.68, NA) and female patient was 3.61 years (95% CI = 2.48, 4.46). The survival time was significant different (p = 0.005)

```{r}
gender_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~gender,
                      data = WHASDS2,
                      type = "kaplan-meier")
gender_KM
```

### KM Plot

KM Plot for gender was shown below

```{r}
ggsurvplot(gender_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```


### Cox Proportional Hazard

If the patient was female, the hazard to die changes by a factor of 1.46 (95% CI = 1.12, 1.92, p < 0.001) compared to male patients. 

```{r}
gender_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~gender,
                 data = WHASDS2)
tidy(gender_CPH)
tidy(gender_CPH, exponentiate =  T)
```


## Univariable - BMI

### Cox Proportional Hazard

For each 1 unit increase of BMI, the log hazard to die changes by a factor of -0.10 (95% = -0.13, -0.07, p < 0.001).

```{r}
BMI_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~bmi,
                 data = WHASDS2)
tidy(BMI_CPH)
tidy(BMI_CPH, exponentiate = T)
```




## Univariable - Systolic BP

### Cox Proportional Hazard

For each 1 mmHg increase of Systolic BP, the log hazard to die changes by a factor of -0.005 (95% = - 0.008, - 0.0001, p = 0.042).

```{r}
SBP_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~sysbp,
                 data = WHASDS2)
tidy(SBP_CPH)
tidy(SBP_CPH, exponentiate = T)
```


## Univariable - Type of MI Order

### KM Estimate

The median survival time for first MI episode was 5.92 years (95% CI = 5.28, NA) and for recurrent MI episode was 3.38 years (95% CI = 2.33, 4.33). The survival time was significant different (p = 0.002)


```{r}
MIOrder_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~MIOrder,
                      data = WHASDS2,
                      type = "kaplan-meier")
MIOrder_KM
```

### KM Plot

KM Plot for type of MI order was shown below

```{r}
ggsurvplot(MIOrder_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Cox Proportional Hazard

If the episode of MI was recurrent episod, the hazard to die changes by a factor of 1.53 (95% CI = 1.17, 2.01, p < 0.001) compared to first MI episode. 

```{r}
MIOrder_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~MIOrder,
                 data = WHASDS2)
tidy(MIOrder_CPH)
tidy(MIOrder_CPH, exponentiate = T)
```

## Univariable - Type of MI

### KM Estimate

The median survival time for Non Q-wave MI was 3.50 years (95% CI = 2.89, 4.46) and Q-wave MI was 6.44 years (95% CI = 6.44, NA). The survival time was significant different (p = 0.005)


```{r}
MIType_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~MIType,
                      data = WHASDS2,
                      type = "kaplan-meier")
MIType_KM
```

### KM Plot

```{r}
ggsurvplot(MIType_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Cox Proportional Hazard

If the MI was Q-wave type, the hazard to die changes by a factor of 0.52 (95% CI = 0.37, 0.72, p < 0.001) compared to non Q-wave type. 

```{r}
MIType_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~MIType,
                 data = WHASDS2)
tidy(MIType_CPH)
tidy(MIType_CPH, exponentiate = T)
```


## Univariable Summary

All variable was significant (at p < 0.05) in simple cox proportional hazard. The univariable analysis result was shown below.


```{r}
SCPH_Result <- matrix(c("Age", 0.07, "(0.05, 0.08)", "< 0.001",
                        "Gender (Female)", 0.38, " (0.11, 0.65)", 0.006,
                        "Systolic BP", -0.01, "(-0.01, 0.00)", 0.042,
                        "BMI (Numerical)", -0.10, "(-0.13, -0.07)", "<0.001",
                       "Type of MI Order (Recurrent)", 0.43, "(0.15, 0.70)", 0.002,
                        "Type of MI (Q-wave)", -0.66, "(-0.99, -0.33)", "<0.001"),
                      ncol = 4,
                      byrow = T)
colnames(SCPH_Result) <- c("Variables", "beta", "(95% CI)", "p-value")
kable(SCPH_Result)
```



# Multivariable Analysis

For multivariable analysis (multiple cox proportional hazard model), all variable with significant simple cox proportional hazard (at p < 0.05) were included in main effect models. The variables are age, gender, systolic BP, BMI, MI order and type of MI. 


## Model 1

For Main Effect Model 1, variables included were age, gender, sbp, bmi, MI order and type of MI. The Cox Proportional Hazard model for Main Effect Model 1 was shown below.

### Cox Propotional Hazard


```{r}
all_CPH <- coxph(Surv(time = dur_year, 
                      FUpStatus == "dead") ~ age + gender + sysbp + bmi + MIOrder + MIType, 
                 data = WHASDS2)
all_CPH
summary(all_CPH)
tidy(all_CPH)
```

## Model 2

For Main Effect Model 2, variable gender were excluded from model 1. The Cox Proportional Hazard Model for Main Effect Model 2 was shown below.

```{r}
AllnoGender_CPH <- coxph(Surv(time = dur_year, 
                      FUpStatus == "dead") ~ age + bmi + sysbp + MIOrder + MIType, 
                 data = WHASDS2)
AllnoGender_CPH
summary(AllnoGender_CPH)
tidy(AllnoGender_CPH)
```



## Model 3 - remove gender & miorder

For Main Effect Model 3, variable gender and MI order were excluded from model 1. The Cox Proportional Hazard Model for Main Effect Model 3 was shown below.

```{r}
Model3_CPH <- coxph(Surv(time = dur_year, 
                      FUpStatus == "dead") ~ age + bmi + sysbp + MIType, 
                 data = WHASDS2)
Model3_CPH
summary(Model3_CPH)
tidy(Model3_CPH)
```


## Model comparison

All the three main effect models were compared. The comparisons were shown below.

```{r}
anova(all_CPH, AllnoGender_CPH, test = "Chisq")
```

```{r}
anova(AllnoGender_CPH, Model3_CPH, test = "Chisq")
```

```{r}
anova(all_CPH, Model3_CPH, test = "Chisq")
```

Since there was no significant different between the three models, model 3 was selected, as it's most parsimonous. The variable includes age, systolic BP, BMI, and type of MI.

```{r}
PrelimFinalModel <- Model3_CPH
```


## Interaction

In this study, three interaction terms will be tested, including

- age and systolic BP
- BMI and Systolic BP
- Systolic BP and Type of MI

### Interaction term between age and systolic BP

There was no significant interaction term between age and systolic BP.

```{r}
PFModel_IntAgeSBP <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~ age + bmi + sysbp + MIType + age:sysbp, data = WHASDS2)
summary(PFModel_IntAgeSBP)
```

### Interaction term between BMI and Systolic BP

There was no significant interaction term between BMI and systolic BP.

```{r}
PFModel_IntBMISBP <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~ age + bmi + sysbp + MIType + bmi:sysbp, data = WHASDS2)
summary(PFModel_IntBMISBP)
```


### Interaction term between Systolic BP and Type of MI

There was no significant interaction term between systolic BP and type of MI.

```{r}
PFModel_IntSBPMIT <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~ age + bmi + sysbp + MIType + sysbp:MIType, data = WHASDS2)
summary(PFModel_IntSBPMIT)
```

# Final Model and Model Checking

Final Model include age, BMI, systolic BP and type of MI variables.

```{r}
FinalModel <- PrelimFinalModel
summary(FinalModel)
```


## Model checking - Linearity

Linearity assumption was tested with Martingale Residual checking. The Martingale Residual Checking show that the model had acceptable linearity, as shown below.


```{r Linearity in hazard assumption}

ggcoxfunctional (Surv(time = dur_year, event = FUpStatus == 'dead') ~  age + bmi, data = WHASDS2) 

```

## Model Checking - Proportionality

Proportionality assumption was checked and show that the assumption was met. The plots were shown below.

```{r Proportional hazard assumption using KM method}

prop.h <- cox.zph(FinalModel, transform = 'km', global = TRUE)
prop.h
plot(prop.h)
```

## Final Model Interpretation

```{r}
tidy(FinalModel)
```

- The final model include age, BMI, systolic BP and type of MI
- For each increment of 1 year of age, the log hazard to die change by 0.06 (95% CI = 0.05, 0.07, p < 0.001) when adjusted to BMI, systolic BP and type of MI
- For each increment of 1 unit of BMI, the log hazard to die change by -0.04 (95% CI = -0.07, -0.01, p = 0.009) when adjusted to age, systolic BP and type of MI
- For each increament of 1 mmHg, the log hazard to die change by -0.004 (95% CI = -0.009, -0.0001, p = 0.046) when adjusted to age, BMI and type of MI


# Prediction

## Create new data

In this study, we would like to predict the survival probablity for data below

- age: 60 and 80 years old
- BMI: 20 and 30
- Systolic BP: 120, 140 and 160 mmHg
- Type of MI: Non Q-wave and Q-wave


## Predict Risk against Average Population

The risk of death for patient (observation 1) with age 60, BMI 20, systolic BP of 120 and with Q-wave MI was 0.67 as compared to average population, while for patient (observation 6) with age 80, BMI 20, systolic BP of 140 with Q-wave MI was 1.94 as compared to average population. The risk of death for other parameters was shown below.


```{r create new data}
new_data <- expand.grid(age =c(60, 80),
                        bmi = c(20,30),
                        sysbp = c(120, 140, 160),
                        MIType = c("Q-wave","non Q-wave"))

risk_new <- predict(FinalModel, newdata = new_data, type = "risk")
head(cbind(new_data, risk_new))
```

## Predict number of event for given follow-up time

We also would like predict the number of event, at 5 and 10 years follow-up time.

For patient (observation 1) with age of 60, BMI 20, systolic BP 120 and q-wave MI, the expected number of event was 0.39 at 5 years follow up. For patient (observation 48) with age of 80, BMI 30, systolic BP 160 and non Q-wave, the expected number of event was 12.03 at 10 years follow up. The expected number of event for other parameters was shown below.

```{r}
new_data2 <- expand.grid(age =c(60, 80),
                        bmi = c(20,30),
                        sysbp = c(120, 140, 160),
                        MIType = c("Q-wave","non Q-wave"),
                        FUpStatus = "dead",
                        dur_year = c(5,10))
predict_new <- predict(FinalModel, newdata = new_data2, type = "expected")
head(cbind(new_data2, predict_new))
tail(cbind(new_data2, predict_new))
```

## Survival Probability

The baseline hazard for our model, at 6 years, was 0.7211136.

```{r}
base_FinalModel <- basehaz(FinalModel)
tail(base_FinalModel)
```

For patient (observation 1) with age 60, BMI 20, systolic BP 120 with q-wave MI, the probability of death was 0.48, while for patient (observation 6) with age 80, BMI 20, systolic BP 140 with q-wave MI, the probability of death was 1.40.

```{r}
Predrisk_new <- 0.7211136*risk_new
head(cbind(new_data, Predrisk_new))
```

