---
title: "Survival Analysis Practical"
output: html_notebook
---

# Preface

## Library

```{r}
library(pacman)
p_load(haven, dplyr, psych, lubridate, knitr, ggplot2, survival, broom, survminer, mfp)
```

## Dataset

Import dataset
```{r}
WHASDS <- read_spss("whas500.sav")
WHASDS
```

Convert to categorical data
```{r}
WHASDS2 <- WHASDS %>%
  mutate(gender = factor(gender,
                         labels = c("male", "female")),
         BMICat = cut(bmi, c(0, 20, 25, 30, 100),
                      labels = c("Underweight", "Normal", "Overweight", "Obese")),
         cvd = factor(cvd,
                      labels = c("No", "Yes")),
         AtrFib = factor(afb,
                         labels = c("No", "Yes")),
         CardioShock = factor(cvd,
                              labels = c("No", "Yes")),
         ConHeartCx = factor(chf,
                             labels = c("No", "Yes")),
         CompHeartBlock = factor(av3,
                                 labels = c("No", "Yes")),
         MIOrder = factor(miord,
                          labels = c("first", "recurrent")),
         MIType = factor(mitype,
                         labels = c("non Q-wave", "Q-wave")),
         DischargeStatus = factor(dstat,
                                  labels = c("alive", "dead")),
         FUpStatus = factor(fstat,
                            labels = c("alive", "dead"))) %>%
  select (id,
          FUpStatus, 
          age,
          gender,
          hr,
          sysbp,
          diasbp,
          bmi,
          BMICat,
          CardioShock,
          ConHeartCx,
          CompHeartBlock,
          MIOrder,
          MIType,
          admitdate,
          disdate,
          fdate,
          DischargeStatus)
WHASDS2
```


calculate duration
```{r}
WHASDS2 <- WHASDS2 %>%
  mutate(dur = as.duration(WHASDS2$admitdate %--% WHASDS2$fdate),
         dur_year = dur/dyears(1)) %>%
  select (id,
          FUpStatus,
          dur_year,
          age,
          gender,
          hr,
          sysbp,
          diasbp,
          bmi,
          BMICat,
          CardioShock,
          ConHeartCx,
          CompHeartBlock,
          MIOrder,
          MIType,
          admitdate,
          disdate,
          fdate,
          DischargeStatus,
          dur)
```


# Descriptive Study

## Summary

summary all data
```{r}
WHASDS2 %>%
  select_if(is.numeric) %>%
  summary()
WHASDS2 %>%
  select_if(is.factor) %>%
  summary()
```

summary all data, group by follow up status **to update**
```{r}
WHASDS2 %>%
  group_by(FUpStatus) %>%
  select_if(is.numeric) %>%
  summary()
WHASDS2 %>%
  group_by(FUpStatus) %>%
  select_if(is.factor) %>%
  summary()
```


## Visualization

barchart for all categorical data
```{r}
ggplot(WHASDS2, aes(x=FUpStatus, fill=FUpStatus)) +
  geom_bar()
ggplot(WHASDS2, aes(x=gender, fill=gender)) +
  geom_bar()
ggplot(WHASDS2, aes(x=BMICat, fill=BMICat)) +
  geom_bar()
ggplot(WHASDS2, aes(x=CardioShock, fill=CardioShock)) +
  geom_bar()
ggplot(WHASDS2, aes(x=ConHeartCx, fill=ConHeartCx)) +
  geom_bar()
ggplot(WHASDS2, aes(x=CompHeartBlock, fill=CompHeartBlock)) +
  geom_bar()
ggplot(WHASDS2, aes(x=MIOrder, fill=MIOrder)) +
  geom_bar()
ggplot(WHASDS2, aes(x=MIType, fill=MIType)) +
  geom_bar()
ggplot(WHASDS2, aes(x=DischargeStatus, fill=DischargeStatus)) +
  geom_bar()
```

histogram for all numerical data
```{r}
ggplot(WHASDS2, aes(dur_year)) +
  geom_histogram(binwidth=0.25, colour="black", fill="white")
ggplot(WHASDS2, aes(age)) +
  geom_histogram(binwidth=2, colour="black", fill="white")
ggplot(WHASDS2, aes(hr)) +
  geom_histogram(binwidth=5, colour="black", fill="white")
ggplot(WHASDS2, aes(sysbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white")
ggplot(WHASDS2, aes(diasbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white")
ggplot(WHASDS2, aes(bmi)) +
  geom_histogram(binwidth=1, colour="black", fill="white")
```

Barchart for all categorical data, divide by follow up status
```{r}
ggplot(WHASDS2, aes(x=gender, fill=gender)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=BMICat, fill=BMICat)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=CardioShock, fill=CardioShock)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=ConHeartCx, fill=ConHeartCx)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=CompHeartBlock, fill=CompHeartBlock)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=MIOrder, fill=MIOrder)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=MIType, fill=MIType)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(x=DischargeStatus, fill=DischargeStatus)) +
  geom_bar() +
  facet_wrap(~FUpStatus)
```

histogram for all numerical data, divide by follow up status
```{r}
ggplot(WHASDS2, aes(age)) +
  geom_histogram(binwidth=2.5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(hr)) +
  geom_histogram(binwidth=5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(sysbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(diasbp)) +
  geom_histogram(binwidth=5, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
ggplot(WHASDS2, aes(bmi)) +
  geom_histogram(binwidth=1, colour="black", fill="white") +
  facet_wrap(~FUpStatus)
```
 
## KM Estimate

```{r}
Overall_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~1,
                      data = WHASDS2,
                      type = "kaplan-meier")
Overall_KM
```

## KM Plot

```{r}
ggsurvplot(Overall_KM,
           data = WHASDS2,
           risk.table = T)
```

## Estimate survival time

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(Overall_KM, probs = c(0.25, 0.5, 0.75))
```

## Survival Probability

```{r}
summary(Overall_KM, times = c(0.5, 1, 3, 5))
```


# Univariable

## Univariable - Age

### Cox Proportional Hazard

```{r}
age_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~age,
                 data = WHASDS2)
age_CPH
tidy(age_CPH)
```

## Univariable - Gender

### KM Estimate

```{r}
gender_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~gender,
                      data = WHASDS2,
                      type = "kaplan-meier")
gender_KM
```

### KM Plot

```{r}
ggsurvplot(gender_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
gender_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~gender,
                      data = WHASDS2,
                      rho = 0)
gender_LR
```

### Cox Proportional Hazard

```{r}
gender_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~gender,
                 data = WHASDS2)
gender_CPH
tidy(gender_CPH)
```


### Survival Time Estimation

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(gender_KM, probs = c(0.25, 0.5, 0.75))
```

## Univariable - Heart Rate

### Cox Proportional Hazard

```{r}
hr_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~hr,
                 data = WHASDS2)
hr_CPH
tidy(hr_CPH)
```



## Univariable - Systolic BP

### Cox Proportional Hazard

```{r}
SBP_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~sysbp,
                 data = WHASDS2)
SBP_CPH
tidy(SBP_CPH)
```





## Univariable - Diastolic BP

### Cox Proportional Hazard

```{r}
DBP_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~diasbp,
                 data = WHASDS2)
DBP_CPH
tidy(DBP_CPH)
```





## Univariable - BMI

### Cox Proportional Hazard

```{r}
BMI_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~bmi,
                 data = WHASDS2)
BMI_CPH
tidy(BMI_CPH)
```

## Univariable - BMI (Category)

### KM Estimate

```{r}
BMICat_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~BMICat,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(BMICat_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2,3,4),
           pval = T)
```

### Log Rank Test

```{r}
BMICat_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~BMICat,
                      data = WHASDS2,
                      rho = 0)
BMICat_LR
```


### Cox Proportional Hazard

```{r}
BMICat_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~BMICat,
                 data = WHASDS2)
BMICat_CPH
tidy(BMICat_CPH)
```

### Survival Time Estimation

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(BMICat_KM, probs = c(0.25, 0.5, 0.75))
```


## Univariable - Type of Cardiogenic Shock

### KM Estimate

```{r}
CardioShock_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~CardioShock,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(CardioShock_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
CardioShock_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~CardioShock,
                      data = WHASDS2,
                      rho = 0)
CardioShock_LR
```

### Cox Proportional Hazard

```{r}
CardioShock_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~CardioShock,
                 data = WHASDS2)
CardioShock_CPH
tidy(CardioShock_CPH)
```


### Survival Time Estimation

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(CardioShock_KM, probs = c(0.25, 0.5, 0.75))
```

## Univariable - Type of Congestive Heart Complication

### KM Estimate

```{r}
CHF_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~ConHeartCx,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(CHF_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
CHF_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~ConHeartCx,
                      data = WHASDS2,
                      rho = 0)
CHF_LR
```


### Cox Proportional Hazard

```{r}
CHF_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~ConHeartCx,
                 data = WHASDS2)
CHF_CPH
tidy(CHF_CPH)
```


### Survival Time Estimation at Q1, median and Q3 survival probability

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(CHF_KM, probs = c(0.25, 0.5, 0.75))
```


## Univariable - Type of Complete Heart Block

### KM Estimate

```{r}
CHB_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~CompHeartBlock,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(CHB_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
CHB_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~CompHeartBlock,
                      data = WHASDS2,
                      rho = 0)
CHB_LR
```

### Cox Proportional Hazard

```{r}
CHB_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~CompHeartBlock,
                 data = WHASDS2)
CHB_CPH
tidy(CHB_CPH)
```


### Survival Time Estimation at Q1, median and Q3 survival probability

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(CHB_KM, probs = c(0.25, 0.5, 0.75))
```



## Univariable - Type of MI Order

### KM Estimate

```{r}
MIOrder_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~MIOrder,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(MIOrder_KM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
MIOrder_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~MIOrder,
                      data = WHASDS2,
                      rho = 0)
MIOrder_LR
```


### Cox Proportional Hazard

```{r}
MIOrder_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~MIOrder,
                 data = WHASDS2)
MIOrder_CPH
tidy(MIOrder_CPH)
```


### Survival Time Estimation at Q1, median and Q3 survival probability

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(MIOrder_KM, probs = c(0.25, 0.5, 0.75))
```


## Univariable - Type of Type of MI

### KM Estimate

```{r}
MIType_KM <- survfit(Surv(time = dur_year, FUpStatus == "dead") ~MIType,
                      data = WHASDS2,
                      type = "kaplan-meier")
```

### KM Plot

```{r}
ggsurvplot(MITypeKM,
           data = WHASDS2,
           risk.table = T,
           linetype = c(1,2),
           pval = T)
```

### Log Rank Test

```{r}
MIType_LR <- survdiff(Surv(time = dur_year, FUpStatus == "dead") ~MIType,
                      data = WHASDS2,
                      rho = 0)
MIType_LR
```

### Cox Proportional Hazard

```{r}
MIType_CPH <- coxph(Surv(time = dur_year, FUpStatus == "dead") ~MIType,
                 data = WHASDS2)
MIType_CPH
tidy(MIType_CPH)
```


### Survival Time Estimation at Q1, median and Q3 survival probability

estimating survival time, at 25th percentile survival probability, median survival probability and 75th percentile survival probability

```{r}
quantile(MIType_KM, probs = c(0.25, 0.5, 0.75))
```


## Univariable Summary

```{r}
SCPH_Result <- matrix(c("Age", 0.07, "(0.05, 0.08)", "< 0.001",
                        "Gender", 0.38, " (0.11, 0.65)", 0.006,
                        "Heart Rate", 0.02, "(0.01, 0.02", "<0.001",
                        "Systolic BP", -0.01, "(-0.01, 0.00)", 0.042,
                        "Diastolic BP", -0.10, "(-0.13, -0.07)", "<0.001",
                        "BMI (Numerical)", -0.10, "(-0.13, -0.07)", "<0.001",
                        "BMI - Underweight", 0, 1, "",
                        "BMI - Normal", -0.76, "(-1.15, -0.48)", "<0.001",
                        "BMI - Overweight", -1.32, "(-1.71, -0.92)", "<0.001",
                        "BMI - Obese", -1.62, "(-2.09,-1.15)", "<0.001",
                        "Type of Cardiogenic Shock - Yes", 0.28, "(-0.05, 0.61)", 0.092,
                        "Type of Congestive Heart Complication - Yes", 1.20, "(0.93, 1.47)", "<0.001",
                        "Type of Complete Heart Block - Yes", 0.48, "(-0.29, 1.24)", 0.219,
                        "Type of MI Order - Recurrent", 0.43, "(0.15, 0.70)", 0.002,
                        "Typw of MI - Q-wave", -0.66, "(-0.99, -0.33)", "<0.001"),
                      ncol = 4,
                      byrow = T)
colnames(SCPH_Result) <- c("Variables", "beta", "(95% CI)", "p-value")
kable(SCPH_Result)
```



# Multivariable

## Model 1 - all variables

IV - age, gender, hr, sbp, dbp, bmi (num), cardiogenic shock, congestive heart complication, complete heart block, MI order and type of MI

### Cox Propotional Hazard


```{r}
all_CPH <- coxph(Surv(time = dur_year, 
                      FUpStatus == "dead") ~ age + gender + hr + sysbp + diasbp + bmi + CardioShock + ConHeartCx + CompHeartBlock + MIOrder + MIType, 
                 data = WHASDS2)
all_CPH
summary(all_CPH)
tidy(all_CPH)
```

## Model 2 - p < 0.25

= all

## Model 3 - p < 0.05

IV - age, gender, hr, sbp, dbp, bmi (num), congestive heart complication, MI order and type of MI

```{r}
sig.05_CPH <- coxph(Surv(time = dur_year, 
                      FUpStatus == "dead") ~ age + gender + hr + sysbp + diasbp + bmi + ConHeartCx + MIOrder + MIType, 
                 data = WHASDS2)
sig.05_CPH
summary(sig.05_CPH)
tidy(sig.05_CPH)
```


## Model comparison

## Interaction

# Final Model and Model Checking

## Propotionality

## Residual

### Residual Plot

# Prediction

## LP Method

## Risk Score Method

(against average sample's parameter)

## Expected Method

# Prediction at 10 years